---
title: Find the Factors of a Given Number
subtitle: "HEllo"
author: ''
date: '2022-11-12'
slug: ''
categories:
  - R
tags:
  - Algorithms
---


## Introduction

Often times simple problems demand harder thought for an elegant solution. 

Here we would like to write an algorithm that returns all the factors for a particular n.  

To optimise our algorithm, let us consider the limits of the factors of a number. Once we reach the square root of $n$ all succesive numbers that are factors of $n$ will have some multiple that belongs to the set of numbers below $n$. Thus, The greatest factor of $n$ has to be $n$ divided by one of the values retrieved from the first array.      . 

We break our solution down into two parts. First, we limit our search for factors below $\sqrt n$ and select  factors of $n$ from this array.Then we search for factors above this threshold, knowing that there exists part of a pair in the array above that successfully multiplies by a number(s) in the current list to yield $n$. The combined result of these two operations will yield all factors of for the given $n$.

 
The first step improves the computation time over the naive approach drastically. $O(\sqrt n)$ from $O(n)$. The second, allows us to shorten the set of numbers from $\sqrt n$ to $n$ we need to divide by.       




### Proposed Algorithm



```{r}

n = 50

factors = function(n){
  
  n_l = 2 : ceiling(sqrt(n)) # initialise range below sqrt n
  f_b = n_l[n %% n_l == 0] # retain factors
  
  if(!length(f_b)) return(c(1,n))  # if prime 

  f_a = c()
  
  for(i in 1: length(f_b)){
    
    upr = ceiling(sqrt(n) + 1):ceiling((n / f_b[i])) # initialise range above sqrt n
    f_a[i] =  upr[upr * f_b[i] == n]
    
  }
  return(sort(c(1,f_b,f_a)))
}

system.time(factors(1e8))

```


We need to loop through each value of `f_b` , as this can't (in any simple way) be done through a vectorised approach. I would be extremely interested to see a fully vectorised implementation. Do get in touch if you're on to something.  


### Naive Approach

```{r naive}

dummy = function(n){
   
  f = 2
  
  for(i in 2:(n/f)){
    if(n %% i == 0) f = i  
   }
}

system.time(dummy(1e8))
```
